Cyclone: High Availability for Persistent Key Value Stores

### 2017-12-23 paper/paper.tex
Updates.

\begin{abstract}
\vspace{0.05in}
Persistent key value stores are an important component of many
distributed data serving solutions with innovations targeted at taking
advantage of growing flash speeds. Unfortunately their performance is
hampered by the need to maintain and replicate a write ahead log to
guarantee availability in the face of machine and storage
failures. Cyclone is a replicated log plug-in for key value stores
that systematically addresses various sources of this bottleneck. It
uses a small amount of non-volatile memory directly addressable by the
CPU - such as in the form of NVDIMMs or Intel 3DXPoint - to remove
block oriented IO devices such as SSDs from the critical path for
appending to the log. This then enables it to address network
overheads using an implementation of the RAFT consensus protocol that
is designed around a userspace network stack to relieve the CPU of the
burden of data copies.  Finally, it provides a way to exploit the
parallelism available in commodity NICs.  Cyclone is able to replicate
millions of small updates per second using only commodity 10 gigabit
ethernet adapters. As a practical application, we use it to improve
the performance (and availability) of RocksDB, a popular persistent
key value store by an order of magnitude when compared to its own
write ahead log without replication.
\end{abstract}

\section{Introduction}
Persistent key value stores are an important component of datacenter scale
storage services. Key value stores such as Rocksdb~\cite{rocksdb},
LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent significant efforts on
both engineering and research fronts. These key value stores include
sophisticated in-memory data structures built around Log Structured Merge (LSM)
trees~\cite{lsmtree} and are heavily tuned to extract maximum performance from
flash-based solid state drives (SSDs).

@@

The second key contribution of this work is in showing that the
throughput bottlenecks in the network stack can be mitigated entirely
in software, once block IO is removed from the critical path. Existing
work on log replication makes the assumption that the bottleneck to
good throughput on the network is the CPU and therefore one must
either resort to network offload mechanisms such as RDMA~\cite{dare,
  farm, faast} or offload the protocol entirely to an
FPGA~\cite{consensus_box}. We demonstrate in this paper that this is
not the case. Rather, once persistence is provided by directly
attached non-volatile memory, thereby eliminating block IO from the
persistence step in replication protocols, the problem becomes akin to
multicasting the same log entry from the leader to all followers. We
design Cyclone's network stack around well known principles in the
software packet switching community. In particular, a careful
implementation of a consensus protocol that relieves the CPU of the
responsibility of data movement permits a high performance
implementation entirely in software that addresses the order of
magnitude network related performance gap in Figure~\ref{fig:problem}
using only commodity 10 Gigabit ethernet.

Both the storage (NVM, SSD) as well as the NIC expose significant
parallelism that is left unused when Cyclone is used to replicate a
single log. The third key contribution of this work is to show how one
can exploit this parallelism for replicating the write ahead log in
key value stores.  Cyclone automatically creates and manages multiple
logs, mapping them to the available parallelism in the system via
multiple instances of the same consensus protocol.

@@

Key value stores such as RocksDB also support atomic writes ({\tt PUT}
operations) to multiple keys, called batched writes, and the ability to take
snapshots.

@@

Cyclone uses the RAFT consensus protocol~\cite{raft} to keep the logs
in sync across the replicas. On failure of the leader replica, a new
leader is automatically elected -- RAFT ensures that the new leader
has the most up to date log. The Cyclone client library automatically
locates the new leader on a failover, resulting in only a brief
interruption in service for clients.  Cyclone guarantees that the
order of operations to any particular key are linearizable. A key sees
the exact same sequence of update operations (including deletes) on
all replicas. These per-key guarantees follow immediately from the
properties of the RAFT log and by requiring clients reading a key to
perform a quorum read (logged as an entry in the RAFT log).

In addition, Cyclone also provides allows clients to do a weak read
that returns the current value of the key at the last known leader
replica. Weak reads can see a possibly stale value
(i.e. is not linearizable) but is howevere guaranteed to have seen all
of its own previous writes. On the positive side, weak reads avoid an
extra round trip for quorum consensus.

Finally, Cyclone allows operations to multiple keys in an atomic
batch, where either all updates in the batch are applied or no
operation in the batch is applied - a property that holds \emph{at all
  replicas}, regardless of failures. Batched operations are supported
in Rocksdb to allow referential integrity between keys - where the
value associated with a key is a reference to another key. We also use
a lightweight batched operation to support snapshots - that can be
viewed as an atomic no-op to all keys. We discuss how batched
operations are supported by Cyclone later in the paper.

@@

\section{Network}
\label{sec:network}
Cyclone uses the RAFT~\cite{raft} consensus protocol to replicate the
log of operations across machines. The bottleneck for log replication
in Figure~\ref{fig:problem} is the CPU - the achieved network
throughput at load is well under the capacity of the 10 Gbps link. The
key to improving CPU utilization for log replication is the insight
that log replcation protocols such as RAFT are oblivious to the actual
contents of the log entry being replicated.  This observation allowed
us to greatly improve replication performance by building a (near)
zero copy implementation of RAFT. Cyclone's implementation of RAFT
does only a small \emph{constant} amount of work per replicated log
entry on the CPU to run the RAFT protocol state machine.  Copying work
that proportional to the amount of data being replicated is done by
the (commodity) NIC.  We therefore shift the bottleneck to the NIC,
making full use of the available network link.

Cyclone provides the guarantee to the client that any
operation responded to is durable and its results \emph{available} as long as a
majority of replicas continue to function. Cyclone's failure model is
\emph{fail-recovery}, where nodes that can recover from failure simply continue
to participate in the protocol. This requires RAFT to persist every log entry
\emph{before} it sends it out for replication and before a follower replica
responds to the leader. Storage overheads are therefore not independent of
network overheads, and we must deal with both storage and networking overheads
simultaneously to improve the performance of replication.

We leverage the two level log from the previous section to remove storage
bottlenecks from the critical path of replication in much the same way we did
for logging on a single machine. RAFT is only aware of and replicates the top
level NVM log. Each replica \emph{independently} drains committed (in terms of
RAFT consensus) entries from the NVM log to the flashlog. Block IO therefore is
no longer a consideration when optimizing data movement over the network, unlike
systems that need to deal with such problems~\cite{reflex}. The task of
replication reduces to that of efficiently sending a block of data already
present in directly attached memory over the network.

The core network operation in replication is to receive a request from the
client at the leader replica and send that exact same request (as a log entry)
to all follower replicas. The leader therefore \emph{multicasts} a received
request packet to follower replicas, a problem well studied in the networking
community when building software packet switches. Software packet switches reach
impressive speeds of millions of packets forwarded per second~\cite{dpdk_perf}, 
a number far in excess of the few thousands of packets we manage in
Figure~\ref{fig:problem}. The key insight in removing network overheads in
Cyclone is therefore to approximate a software packet switch for the networking
component of Cyclone. We do this by implementing Cyclone's network stack on top
of the Data Plane Development Kit (DPDK)~\cite{dpdk}, that provides low latency
userspace access to an ethernet NIC. We discuss Cyclone's network protocol in
Section~\ref{sec:netprot}.

However, the flow of packets through a software packet switch is usually very
simple: the packet enters through an ethernet port and after a simple (usually
stateless) decision is sent out through a set of chosen ports.  In contrast, log
replication requires a complex protocol state machine to decide what to do with
the packet. In addition, unlike software switches we cannot simply forget a
packet after transmitting it. We must append it to the log and act on it after
consensus is reached by a majority quorum. In Section~\ref{sec:dm} we cover the
techniques we use in Cyclone to ameliorate these overheads.

\subsection{DPDK}
\label{sec:netprot}
The Data Plane Development Kit~\cite{dpdk} provides low latency userspace access
to an ethernet NIC, permitting the application to directly send and receive raw
ethernet frames via the transmit and receive queues on the NIC. DPDK is often
used by developers of software packet switches and therefore we leverage
DPDK as a library for building Cyclone.

DPDK does not by itself provide a TCP stack. But this is not a problem, since
RAFT (and indeed most consensus protocols) tolerate network losses and
reordering by design due to the need to support asynchronous communication. In
addition, most datacenter networks rarely drop or reorder packets and provide
full bisection bandwidth between servers that might serve as Cyclone replicas. We
therefore jettisoned TCP and chose to send raw IP packets encapsulated in
ethernet frames. Cyclone takes complete control of an ethernet interface,
receiving all packets directed to it, while the IP and ethernet addresses
provide sufficient information to route the packet if necessary through multiple
switches. We currently follow this communication model both for server to server
communication between replicas as well as client to server communication for
requests and responses. Although a detailed evaluation is made later, switching
from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
testbed from {\tt 18 us} to as low as {\tt 5 us}, providing a significant boost
to performance.

\subsection{Addressing Overheads}
\label{sec:dm}
Software packet switches built on DPDK try to touch as little of the packet data
as possible, minimizing movement of data up and down the cache and memory
hierarchy. Log replication looks like multicast that does not require deep
packet inspection. Software packet switches implementing multicast therefore
simply manipulate packet headers to produce new packets to send on the output
ports. We designed our implementation of the RAFT protocol based on the same
principle.

@@

\begin{figure}
\centering
\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
\caption{NVM log structure}
\label{fig:nvm_log}
\end{figure}

The pseudocode in Figure~\ref{fig:control_plane} describes part of the packet
handling code in Cyclone organized as event handlers triggered on receiving a
packet at the leader. We focus on only one key event for brevity: the event
where a request is received from a client. The first step is to check that this
replica is indeed the RAFT leader in the current view (term). If not, the
message is simply dropped (a timeout causes the client to try a different
server). It then prepends RAFT related information to the packet - this includes
the current term and log index for this entry. Next, it appends a pointer to the
log in Figure~\ref{fig:nvm_log} effectively appending the packet to the
persistent log. Finally, it transmits the packet to follower replicas.

@@

We direct DPDK to use pages backed by NVM for packet buffers. DPDK uses a
concurrent memory allocator based on reference counting - used by both the NIC
driver and CPU cores. This means that packet allocation and deallocation does
not happen in the same order as their corresponding position in the replicated
RAFT log. To deal with this, we use a level of indirection as shown in
Figure~\ref{fig:nvm_log}. The NVM log is maintained as a circular log of fixed
sized pointers to the actual packets. Adding a level of indirection in the NVM
log allows us to separate the FIFO ordered circular log being manipulated by
RAFT from packet data being managed by the memory allocator of DPDK. Both the
circular log and packet data are in NVM. An advantage of this scheme is that it
makes recovery from NVM easy -- appends to the circular pointer log are atomic
and we can use the pointer log to recover allocator state i.e. what pieces of
NVM are currently in use by the log.

RAFT requires that the log entry be persisted \emph{before} it is multicast out
to follower replicas. DPDK userspace NIC drivers operate in DDIO
mode~\cite{ddio} where the packet is directly written into the CPU cache rather
than first being DMAed into DRAM and then fetched by the CPU on demand. We need
to persist the packet to directly attached NVM by executing a cacheline flush
({\tt clflush}) instruction for every cacheline in the packet and the pointer in
the pointer buffer to persist these via the memory bus. This is not too onerous
a burden because we can use the newly introduced {\tt
  clflush-opt}~\cite{clflush_opt} instruction specifically intended to
efficiently flush to persistent memory without the overhead of the serialization
normally introduced by {\tt clflush}. This allows us to hit full memory
bandwidth on current generation platforms, a quantity in excess of 200 Gb/s per
core which is well above the near term speeds of commodity NICs. We execute a
single serializing {\tt sfence} before the sequence of {\tt clflushes} to make
sure any dirty cachelines due to header manipulation related writes from the CPU
are sent to cache.

Although RAFT is an efficient consensus protocol in the common case, the
protocol state machine still adds significant overhead to each packet, relative
to the time for the packet to flow in from the NIC and back out to the replicas.
We address this problem for the \emph{loaded} case using batching - treating a
whole sequence of client commands as a single RAFT log entry, while avoiding any
copies to group these packets together. Figure~\ref{fig:batching} illustrates
how this is done. We use a burst receive call available in the DPDK userspace
driver to receive a burst of client packets at a time. We then chain these
packets together and treat them as a single log entry from the perspective of
RAFT, amortizing the control plane overheads over the packets (at most 32 at a
time due to current driver limitations). Crucially, batching in Cyclone does not
involve a latency-throughput tradeoff like in many other
systems~\cite{ix-dataplane}. The batch receive call we use in DPDK returns
immediately with whatever number of packets is available, including zero. We
always flush the transmit buffer after every call to DPDK to transmit packets to
replicas. Therefore, we never tradeoff latency for throughput when batching. We
also receive log entries in batches at follower replicas and return a single
acknowledgment for the entire batch, speeding up the progress of the protocol.

\begin{figure}
  \centering
  \includegraphics[scale=0.3]{figures2/batching.pdf}
  \caption{Batching}
  \label{fig:batching}
\end{figure}

Our choice of RAFT as a consensus protocol is driven by the need to efficiently
maintain the log datastructure. Unlike alternatives such as
MultiPaxos~\cite{multipaxos} or Quorum based replication~\cite{quorum}, the
leader is guaranteed to have the most up to date logs. There is therefore no
case where a leader needs to receive log entries to ``fill'' holes in its logs
from follower replicas, simplifying the protocol state machine we need to
implement. A leader can immediately reject any responses that are not at least
as current as its view (term). In turn, the persistent log operates as a double
ended queue, with entries either being appended (or possibly deleted at
followers) at the end and being deleted from the front (after commit). This is
critical to efficient operation of Cyclone, as it allows a simple top level
structure (a circular buffer of pointers) to represent the log.

>>>>

\begin{abstract}
\vspace{0.05in}
Persistent key value stores are an important component of many
distributed data serving solutions with innovations targeted at taking
advantage of growing flash speeds. Unfortunately their performance is
hampered by the need to maintain and replicate a write ahead log to
guarantee availability in the face of machine and storage
failures. Cyclone is a replicated log plug-in for key value stores
that systematically addresses various sources of this bottleneck. It
uses a small amount of non-volatile memory directly addressable by the
CPU - such as in the form of NVDIMMs or Intel 3DXPoint - to remove
block oriented IO devices such as SSDs from the critical path for
appending to the log. This then enables it to address network
overheads using an implementation of the RAFT consensus protocol that
is designed around a userspace network stack to relieve the CPU of the
burden of data copies.  Finally, it provides a way to exploit the
parallelism available in commodity NICs.  Cyclone is able to replicate
millions of small updates per second using only commodity 10 gigabit
ethernet adapters. As a practical application, we use it to improve
the performance of RocksDB, a popular persistent key value store by an
order of magnitude when compared to its own write ahead log without
replication.
\end{abstract}

\section{Introduction}
Persistent key value stores are an important component of datacenter
scale storage services. Key value stores such as
Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb}
represent significant efforts on both engineering and research
fronts. They include sophisticated in-memory data structures built
around Log Structured Merge (LSM) trees~\cite{lsmtree} and are heavily
tuned to extract maximum performance from flash-based solid state
drives (SSDs).

@@

The second key contribution of this work is in showing that the
throughput bottlenecks in the network stack can be mitigated entirely
in software. Existing work on log replication makes the assumption
that the bottleneck to good throughput on the network is the CPU and
therefore one must either resort to network offload mechanisms such as
RDMA~\cite{dare, farm, faast} or offload the protocol entirely to an
FPGA~\cite{consensus_box}. We demonstrate in this paper that this is
not the case. In particular, a careful implementation of a consensus
protocol that relieves the CPU of the responsibility of data movement
permits a high performance implementation entirely in software that
addresses the order of magnitude network related performance gap in
Figure~\ref{fig:problem} using only commodity 10 Gigabit ethernet.

Commodity network interfaces expose significant parallelism via
multiple transmit receive queues that is left unused when Cyclone is
used to replicate a single log. The third key contribution of this
work is to show how one can exploit this parallelism for replicating
the write ahead log in key value stores. Cyclone creates and manages
multiple logs, mapping them to the available parallelism via multiple
instances of the same consensus protocol.

@@

Key value stores such as RocksDB also support atomic writes ({\tt PUT}
operations) to multiple keys, called batched writes, and the ability
to take snapshots. Batched operations are supported in Rocksdb to
allow referential integrity between keys - where the value associated
with a key is a reference to another key.

@@

Cyclone uses the RAFT consensus protocol~\cite{raft} to keep the logs
in sync across the replicas. On failure of the leader replica, a new
leader is automatically elected -- RAFT ensures that the new leader
has the most up to date log. The Cyclone client library automatically
locates the new leader on a failover, resulting in only a brief
interruption in service for clients.  Cyclone guarantees that each key 
sees the same linearizable order of operations. (including deletes) on
all replicas. These per-key guarantees follow immediately from the
properties of the RAFT log and by requiring clients reading a key to
perform a quorum read (logged as an entry in the RAFT log).

In addition, Cyclone allows clients to do a weak read that returns the
current value of the key at the last known leader replica. Weak reads
from a client can see a possibly stale value (are not linearizable)
but are guaranteed to have seen all of its own previous writes. On the
positive side, weak reads avoid an extra round trip for quorum
consensus.

For completeness, Cyclone allows operations to multiple keys in an
atomic batch, where either all updates in the batch are applied or no
operation in the batch is applied - a property that holds \emph{at all
  replicas}, regardless of failures.  We also use a lightweight
batched operation to support snapshots - that can be viewed as an
atomic no-op to all keys.

@@

\section{Network}
\label{sec:network}
Cyclone uses the RAFT~\cite{raft} consensus protocol to replicate the
log of operations across machines. The bottleneck for log replication
in Figure~\ref{fig:problem} is the CPU - the achieved network
throughput at load is well under the capacity of the 10 Gbps link. The
key to improving CPU utilization for log replication is the insight
that log replcation protocols such as RAFT are oblivious to the actual
contents of the log entry being replicated.  This observation allowed
us to greatly improve replication performance by building a
zero copy implementation of RAFT. Cyclone's implementation of RAFT
does only a small \emph{constant} amount of work per replicated log
entry on the CPU to run the RAFT protocol state machine.  Copying work
is proportional to the amount of data being replicated and is done by
the (commodity) NIC.  We therefore shift the bottleneck to the NIC,
making full use of the available network link.

RAFT is only aware of and replicates the top level NVM log. Each
replica \emph{independently} drains committed (in terms of RAFT
consensus) entries from the NVM log to the flashlog. Block IO
therefore is no longer a consideration when optimizing data movement
over the network, unlike systems that need to deal with such
problems~\cite{reflex}.

\subsection{DPDK}
\label{sec:netprot}
The Data Plane Development Kit~\cite{dpdk} provides low latency userspace access
to an ethernet NIC, permitting the application to directly send and receive raw
ethernet frames via the transmit and receive queues on the NIC. DPDK is often
used by developers of software packet switches and is designed to allow the CPU 
to make a switching decision by only examining the first few bytes of the packet 
(containing the header).The NIC does all copying work minimizing the memory access 
required from the CPU. This makes it an ideal candidate to build Cyclone.

DPDK does not by itself provide a TCP stack. This is not a problem, since
RAFT (and indeed most consensus protocols) tolerate network losses and
reordering by design due to the need to support asynchronous communication. In
addition, most datacenter networks rarely drop or reorder packets and provide
full bisection bandwidth between servers that might serve as Cyclone replicas. We
therefore jettisoned TCP and chose to send raw IP packets encapsulated in
ethernet frames. Cyclone takes complete control of an ethernet interface,
receiving all packets directed to it, while the IP and ethernet addresses
provide sufficient information to route the packet if necessary through multiple
switches. We currently follow this communication model both for server to server
communication between replicas as well as client to server communication for
requests and responses. Although a detailed evaluation is made later, switching
from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
testbed from {\tt 18 us} to as low as {\tt 5 us}, providing a significant boost
to performance.

\subsection{Addressing Overheads}
\label{sec:dm}
Software packet switches built on DPDK try to touch as little of the packet data
as possible, minimizing movement of data up and down the cache and memory
hierarchy. Software packet switches implementing multicast therefore
simply manipulate packet headers to produce new packets to send on the output
ports. We designed our implementation of the RAFT protocol based on the same
principle.

@@

%\begin{figure}
%\centering
%\includegraphics[scale=0.3]{figures2/nvm_log.pdf}
%\caption{NVM log structure}
%\label{fig:nvm_log}
%\end{figure}

he pseudocode in Figure~\ref{fig:control_plane} describes part of the packet
handling code in Cyclone organized as event handlers triggered on receiving a
packet at the leader. We focus on only one key event for brevity: the event
where a request is received from a client. The first step is to check that this
replica is indeed the RAFT leader in the current view (term). If not, the
message is simply dropped (a timeout causes the client to try a different
server). It then prepends RAFT related information to the packet - this includes
the current term and log index for this entry. Next, it appends a pointer to the packet to 
the NVM log. Finally, it transmits the packet to follower replicas.

@@

The NVM log is maintained as a circular log of fixed sized pointers to
the actual packets. Adding a level of
indirection in the NVM log allows us to separate the FIFO ordered
circular log being manipulated by RAFT from packet data being managed
by the memory allocator of DPDK. We use the pointer log to recover
allocator state i.e. what pieces of NVM are currently in use by the
log.

%DPDK userspace NIC drivers operate in DDIO mode~\cite{ddio} where the
%packet is directly written into the CPU cache rather than first being
%DMAed into DRAM and then fetched by the CPU on demand. We need to
%persist the packet to directly attached NVM by executing a cacheline
%flush ({\tt clflush}) instruction for every cacheline in the packet
%and the pointer in the pointer buffer to persist these via the memory
%bus. This is not too onerous a burden because we can use the newly
%introduced {\tt clflush-opt}~\cite{clflush_opt} instruction
%specifically intended to efficiently flush to persistent memory
%without the overhead of the serialization normally introduced by {\tt
%  clflush}. This allows us to hit full memory bandwidth on current
%generation platforms, a quantity in excess of 200 Gb/s per core which
%is well above the near term speeds of commodity NICs. This departure
%from a pure zero copy implementation of RAFT is necessitated by the
%lack of current platform support for DMA to persistent memory.
%Although RAFT is an efficient consensus protocol in the common case,
%the protocol state machine still adds significant overhead to each
%packet, relative to the time for the packet to flow in from the NIC
%and back out to the replicas.

We further reduce work per-packet for the \emph{loaded} case using
batching - treating a whole sequence of client commands as a single
RAFT log entry, while avoiding any copies to group these packets
together. We do this by making use of a DPDK feature that allows
chaining received packet mbufs together and treating the chain as a
single packet.  Finally, we pipline the RAFT protocol state machine by
contnuing to transmit log entries to follower replicas without waiting
for acknowledgments. The RAFT protocol allows tracking multiple log
entries in flight via their entry index in the log and this allows us
to keep the link busy without being limited by the round trip time.
At follower replicas we also receive packets containing log entries in
batches and send back a single acknowledgment for the whole batch.

Our choice of RAFT as a consensus protocol is driven by the need to
efficiently maintain the log datastructure. Unlike alternatives such
as MultiPaxos~\cite{multipaxos} or Quorum based
replication~\cite{quorum}, the leader is guaranteed to have the most
up to date logs. There is therefore no case where a leader needs to
receive log entries to ``fill'' holes in its logs from follower
replicas, simplifying the protocol state machine we need to implement
and permitting further optimization.
